---
title: "Basics of image data and spatial patterns analysis in R EBImage"
author: "Andrzej Oleś and Wolfgang Huber"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  ioslides_presentation:
    fig_retina: null
    css: style.css
    widescreen: false
bibliography: BioC2015Oles.bib
vignette: >
  %\VignetteIndexEntry{Basics of image data and spatial patterns analysis in EBImage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r echo=FALSE, message=FALSE}
library(EBImage)
f = system.file("images", "sample.png", package="EBImage")
img = readImage(f)
.dim = dim(img)
.dpi = 100

library(knitr)
opts_chunk$set(error=FALSE)
```

## EBImage

Image processing and analysis toolbox for *R*

- Reading and writing of image files
- Interactive image viewer
- Image manipulation, transformation and filtering
- Object detection and feature extraction

Since *Bioconductor* 1.8 (2006)


<table width="80%" style="font-size: 80%">
<tr><td>
<h3>Original developers:</h3>
Oleg Sklyar<br/>
Wolfgang Huber<br/>
Mike Smith<br/>
Gregoire Pau
</td><td>
<h3>Contributors:</h3>
Joseph Barry<br/>
Bernd Fischer<br/>
Ilia Kats<br/>
Philip A. Marais<br/>
</td></tr>
</table>

## Reading and displaying images

```{r, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
library(EBImage)

f = system.file("images", "sample.png", package="EBImage")
img = readImage(f)
display(img)
```

```{r, eval=FALSE}
bioc = readImage("http://www.bioconductor.org/images/logo/jpg/bioconductor_logo_rgb.jpg")
display(bioc)
```

## Adding text labels

```{r, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2, results='hide'}
display(img, method = "raster")
text(x = 20, y = 20, label = "Parrots", adj = c(0,1), col = "orange", cex = 2)

filename = "parrots.jpg"
dev.print(jpeg, filename = filename , width = dim(img)[1], height = dim(img)[2])
```{r}
file.info(filename)$size
```

## Writing images

Supported file formats: JPEG, PNG and TIFF.

```{r}
writeImage(img, "sample.jpeg", quality = 85)

writeImage(img, "sample.tiff")
writeImage(img, "sample_compressed.tiff", compression = "deflate")

files = list.files(pattern = "sample*")
data.frame(row.names=files, size=file.size(files))
```

## Image representation

<table width="100%">
<tr><td style="vertical-align: top;">
Multi-dimensional pixel intensity arrays

<div style="font-size:75%">
- (x, y)   
- (x, y, **z**}) z-stack
- (x, y, **t**) time-lapse
- (x, y, **c**) channels
- (x, y, c, z, t, ...)
</div>

</td><td>
<div style="width:360px;">![sample](images/sample.svg)</div>
</td></tr>
<tr><td style="vertical-align: top;">
<div style="width:250px;">![replicates](images/replicates.svg)</div>
</td><td>
![sample-rgb](images/sample-rgb.svg)
</td></tr>
</table>

## Image representation

```{r}
str(img)
dim(img)
imageData(img)[1:3, 1:6]
```

## Image summary

```{r}
img
```

## Image histogram

```{r, fig.width=5, fig.height=5}
hist(img)
```

## Color images

```{r, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2,}
f = system.file("images", "sample-color.png", package="EBImage")
imgcol = readImage(f)
display(imgcol)
print(imgcol, short = TRUE)
```

## Image stacks

```{r, echo=FALSE}
nuc = readImage(system.file("images", "nuclei.tif", package="EBImage"))
```{r, fig.width=dim(nuc)[1L]/.dpi, fig.height=dim(nuc)[2L]/.dpi, dpi=.dpi/2}
nuc = readImage(system.file("images", "nuclei.tif", package="EBImage"))
print(nuc, short = TRUE)
display(nuc)
```

## Image stacks

```{r, fig.width=dim(nuc)[1L]/.dpi, fig.height=dim(nuc)[2L]/.dpi, dpi=.dpi}
display(nuc, method = "raster", all = TRUE)
```

## Manipulating images

Being numeric arrays, images can be conveniently manipulated by any of R's arithmetic operators.

### Cropping 

```{r fig.width=384L/.dpi, fig.height=384L/.dpi, dpi=.dpi/2}
img = img[366:749, 58:441]
```

### Inversion

```{r negative, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
img_neg = max(img) - img
display( img_neg )
```

## Manipulating images

### Brightness, contrast, and gamma correction

```{r arithmetic, fig.width=4*dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
img_comb = combine(
  img,
  img + 0.3,
  img * 2,
  img ^ 0.5
)

display( tile(img_comb, numberOfFrames(img_comb), lwd=10, fg.col = "white") )
```

## Binary images

```{r, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
( img_thresh = img > .5 )
display( img_thresh )
```

## Spatial transformations
### Translation

```{r translate, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
img_translate = translate(img, c(100,-50))
display(img_translate)
```

## Spatial transformations
### Rotation

```{r rotate-pre, echo=FALSE}
img_rotate = rotate(img, 30, bg.col = "white")
```{r rotate, eval=FALSE}
img_rotate = rotate(img, 30, bg.col = "white")
display(img_rotate)
```{r rotate-post, echo=FALSE, fig.width=dim(img_rotate)[1L]/.dpi, fig.height=dim(img_rotate)[2L]/.dpi, dpi=.dpi/2}
display(img_rotate)
```

## Spatial transformations
### Scaling

```{r resize-pre, echo=FALSE}
img_resize = resize(img, w=512, h=256)
```{r resize, eval=FALSE}
img_resize = resize(img, w=512, h=256)
display(img_resize )
```{r resize-post, echo=FALSE, fig.width=dim(img_resize)[1L]/.dpi, fig.height=dim(img_resize)[2L]/.dpi, dpi=.dpi/2}
display(img_resize)
```

## Spatial transformations
### Vertical and horizontal reflection

```{r flipflop, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
display( flip(img) )
display( flop(img) )
```


## Affine transformation

Described by a 3x2 transformation matrix $\textbf{m}$

<div style="font-size:75%">
$$\begin{bmatrix} x'_1 & y'_1 \\
x'_2 & y'_2 \\
\vdots & \vdots \\
\end{bmatrix} =
\begin{bmatrix} x_1 & y_1 & 1 \\
x_2 & y_2 & 1 \\
\vdots & \vdots & \ddots \\
\end{bmatrix}
\times \textbf{m}$$
<br/>
$$\textbf{m}_{\text{translation}}=\begin{bmatrix} 1 & 0 \\ 0 & 1 \\ t_x & t_y \\ \end{bmatrix},\quad
\textbf{m}_{\text{rotation}}=\begin{bmatrix} \cos{\theta} & \sin{\theta} \\ -\sin{\theta} & \cos{\theta} \\ 0 & 0 \\ \end{bmatrix},\quad
\textbf{m}_{\text{scaling}}=\begin{bmatrix} W & 0 \\ 0 & H \\ 0 & 0 \\ \end{bmatrix}$$
</div>

## Affine transformation

Example: horizontal sheer mapping

```{r, fig.width=dim(img)[1L]/.dpi, fig.height=dim(img)[2L]/.dpi, dpi=.dpi/2}
m =  matrix(c(1, -.5, 128, 0, 1, 0), nrow=3, ncol=2)
m
img_affine = affine(img, m)
display( img_affine )
```

## Image transposition

```{r transpose, fig.width=dim(imgcol)[2L]/.dpi, fig.height=dim(imgcol)[1L]/.dpi, dpi=.dpi/2}
imgcol_t = transpose(imgcol)
display( imgcol_t )
```

## Tresholding images

Furthermore, we can crop and threshold images with standard matrix operations.

```{r}
img_thresh = img > .5
display(img_thresh)
```{r cropthreshold-post, echo=FALSE, fig.width=dim(img_thresh)[1L]/.dpi, fig.height=dim(img_thresh)[2L]/.dpi, dpi=.dpi/2}
display(img_thresh)
```

Thresholding returns an `Image` object with binarized pixels values. 
```{r img_thresh}
img_thresh
```

