---
title: "Basics of image data and spatial patterns analysis in R"
author: "Andrzej Oleś and Wolfgang Huber"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  ioslides_presentation:
    fig_retina: null
    css: style.css
    widescreen: false
bibliography: BioC2015Oles.bib
vignette: >
  %\VignetteIndexEntry{Basics of image data and spatial patterns analysis in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, message=FALSE}
library(EBImage)
library(knitr)

.dpi = 10
opts_chunk$set(collapse=FALSE, prompt=TRUE, comment=NA)
```

# Spatial statistics: point processes 

Spatial distribution of the position

## Interaction between immune and cancer cells

Lymph nodes

- immunologic filter composed of B and T cells, dendric cells (DC) and macrophages
- clinical significance in cancer staging

<table style="border-spacing: 30px; border-collapse: separate;">
<tr><td>
```{r, echo=FALSE}
.lymphnode = readImage( system.file("images", "testscan_1_2_RGB99-4525D.jpg", package = "BioC2015Oles") )
```{r, echo=FALSE, fig.width=dim(.lymphnode)[1L]/(4*.dpi), fig.height=dim(.lymphnode)[2L]/(4*.dpi), dpi=.dpi}
display(.lymphnode)
```
</td><td style="vertical-align: bottom">
<div style="font-size:80%">
Breast cancer lymph node biopsy [@Setiadi2010].

Cell typing was done by staining the cells with protein antibodies that provide specific signatures.
</div>
</td></tr>
</table>

## Lymph node biopsies

```{r}
data(brcalymphnode, package = "BioC2015Oles")
head(brcalymphnode)
nrow(brcalymphnode)
table(brcalymphnode$class)
```

## Spatial distribution of cells

Plot of the *x* and *y* positions of the T cells (left) and tumor cells (right).

```{r, fig.width=7, fig.height=3.5}
par(mfrow = c(1,2), mar = c(4.5, 4.5, 0.5, 0.5))
xlim = range(brcalymphnode$x)
ylim = range(brcalymphnode$y)
cols = c(`T_cells` = "dodgerblue4", `Tumor` = "darkmagenta")
for(i in seq_along(cols))
  plot(subset(brcalymphnode, class==names(cols)[i])[, c("x", "y")], 
       pch = ".", asp = 1, xlim = xlim, ylim = ylim, col = cols[i])
```

## Analysis of spatial point patterns

Marked spatial point process:

- set of isolated points located in a mathematical space `(xlim, ylim)`
- points marked with certain properties (factor `class`)


```{r, message=FALSE}
library("spatstat")

ln = with(brcalymphnode, 
          ppp(x = x, y = y, marks = class, xrange = xlim, yrange = ylim) )
ln
```

## Convex hull

```{r, fig.width=2, fig.height=2}
chull = convexhull(ln)
par( mar = c(0, 0, 0, 0) )
plot(Window(ln), main = NULL, asp = 1)
plot(chull, lty = 2, col = "lightblue", add = TRUE)
Window(ln) = chull
ln
```

## First order effects: the intensity

Number of points lying in a circle of area $a$ around a point $p=(x,y)$:
$$N(p, a)$$

Local intensity:
<div style="font-size:80%">
$$\lambda(p) = \lim_{a\rightarrow 0} \frac{E[ N(p, a)]}{a},$$

For a stationary process $\lambda(p) = \text{const.}$; then the intensity in an area $A$ is proportional to the area:
$E(N(\cdot, A)) = \lambda A$. 
</div>

## Estimating the intensity

<p>
Kernel smoothed intensity estimate of a point pattern [@Diggle1985]

<div style="font-size:80%">
$$\hat{\lambda}(p) = \sum_i e(p_i) K(p-p_i),$$
where $K$ is the Gaussian smoothing kernel, and $e(p_i)$ is an edge correction factor.
</div>
</p>
```{r density.ppp1, fig.width=9.5, fig.height=2}
par(mfrow = c(1,2), mar = c(0, 0, 0, 2))

plot( density(subset(ln, marks=="Tumor"), edge = TRUE, diggle = TRUE), main = NULL )
plot( density(subset(ln, marks=="Tumor"), edge = FALSE), main = NULL )
```

## Probability of cell classes

Estimate of the spatially-varying probability of a particular event type.

```{r}
plot( relrisk(ln, sigma = 250), main = NULL, mfrow = c(1, 4) )
```

## Second order effects: clustering

Spatial clustering (or anticlustering) of points: whether and to what extent neighbouring points are more/less dense than expected by chance.

The nearest neighbour (NN) distance distribution function $G$

- cumulative distribution function of the distance $W$ from a typical random point to its nearest neigbour.

Homogenous Poisson process

$$G(r) = P(W\leq r) = 1-e^{-\lambda \pi r^2}.$$

## Estimation of the NN distance function *G*

```{r Gest}
gln = Gest(ln)
gln
```   

## Estimation of the NN distance function *G*

```{r, fig.width=5, fig.height=5, message=FALSE}
library(RColorBrewer)

plot(gln, lty = 1, col = brewer.pal(4, "Set1"))
```

- effect of finite cell size
- clustering at large distances

## Ripley's *K* function

The *K* function of a stationary point process is defined so that $\lambda K(r)$ is the expected number of 
additional points within a distance $r$ of a given, randomly picked point. 

Generalization to inhomogenous point processes:

$$K_{\mbox{\scriptsize inhom}}(r)= \sum_{i,j} {\mathbf 1}_{d(p_i, p_j) \le r} \times \frac{e(p_i, p_j, r)} { \lambda(x_i)  \lambda(x_j) },$$

where $d(p_i, p_j)$ is the distance between points $p_i$ and $p_j$, and 
$e(p_i, p_j, r)$ is an edge correction factor.

More useful for estimation and visualisation is the $L$ function:

$$L(r)=\sqrt{\frac{K(r)}{\pi}}.$$

For a Poisson point pattern, the theoretical value is $L(r) = r$. 

## Estimate of the inhomogeneous *L*-function

```{r, eval=FALSE}
Lln = Linhom( subset(ln, marks=="T_cells") )
Lln
```{r, echo=FALSE}
data(Lln, package = "BioC2015Oles")
Lln
```

## Ripley's *K* function

```{r, fig.width=5, fig.height=5}
plot(Lln, lty = 1, col = brewer.pal(3, "Set1"))
```   

## Pair correlation function

```{r, eval=FALSE}
pcfln = pcf( Kinhom(subset(ln, marks=="T_cells")) )
plot(pcfln, lty = 1, log = "x")
```{r, fig.width=5, fig.height=5, echo=FALSE}
data(pcfln, package = "BioC2015Oles")
plot(pcfln, lty = 1, log = "x")
```

## References
